---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Using the Project

## Modify the content

To modify the content just clone the repository, modify the file and commit and push the changes. To modify the actual content of the book (i.e., Chapters) you can modify the `quarto` (`.qmd`) files within `chapters/`. Even if you do not re-render the files you modify the `qmd` files and I will render the project. They are just plain text files. 

## Rendering the book

In addition to file changing, it is possible to re-render the book. However to do this one need to have **Quarto** installed along with all the required packages used within each chapter.

If you want to render the book the best option is to:

- have R and R Studio installed (not mandatory R Studio but easier)
- have Quarto installed
- install the `renv` package within R Studio (`install.packages("renv")`)
- open the `rep3.Rproj` with R Studio

The `renv` package is a way to manage and install all the R packages used in a project and should install all the required packages. 

## Preview the book

The last updated version of the book can be found at: `_book/index.html`. Just opening this file with a standard browser is enough to navigate and see the final result. If the book is not re-rendered the updates to the single `*.qmd` files will not be integrated.

```{r, results = "asis", echo = FALSE}
system("sh quarto-scripts/safe-todoget.sh")
todo <- readLines("todoget.md")
cat(todo, sep = "\n")
```