---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Contents

## Chapter 1 - what is a replication

- theoretical introduction to the concept of replication
- papers presented in the summer school
- important references
- replicability crisis

#### References

- Machery, E. (2020). What is a replication? Philosophy of Science, 87(4), 545–567. https://www.cambridge.org/core/journals/philosophy-of-science/article/what-is-a-replication/F41B751ECC31462CBBD46711BC733AEE
- Miller, J. (2009). What is the probability of replicating a statistically significant effect? Psychonomic Bulletin & Review, 16(4), 617–640. https://doi.org/10.3758/PBR.16.4.617

## Chapter 2 - replicability and reproducibility

- replicability vs reproducibility
- reproducibility tools
- suggested references and tools

## Chapter 3 - formalizing and measuring replication

- the meta-analysis model for uniformity and notation
- problem of measuring replication
- simulation introduction

#### References

- Hedges, L. V., & Schauer, J. M. (2019). Statistical analyses for studying replication: Meta-analytic perspectives. Psychological Methods, 24(5), 557–570. https://doi.org/10.1037/met0000189
- Schauer, J. M. (2022). Replicability and Meta-Analysis. In W. O’Donohue, A. Masuda, & S. Lilienfeld (Eds.), Avoiding Questionable Research Practices in Applied Psychology (pp. 301–342). Springer International Publishing. https://doi.org/10.1007/978-3-031-04968-2_14

## Chapter 4 - frequentists and bayesian methods

- review of the methods with the R application
- pro and cons of method with the power consideration of Hedges
- r packages with all the function for the replications
- review of existing R packages

#### References

- Camerer, C. F., Dreber, A., Holzmeister, F., Ho, T.-H., Huber, J., Johannesson, M., Kirchler, M., Nave, G., Nosek, B. A., Pfeiffer, T., Altmejd, A., Buttrick, N., Chan, T., Chen, Y., Forsell, E., Gampa, A., Heikensten, E., Hummer, L., Imai, T., … Wu, H. (2018). Evaluating the replicability of social science experiments in Nature and Science between 2010 and 2015. Nature Human Behaviour, 2(9), 637–644. https://doi.org/10.1038/s41562-018-0399-z
- Errington, T. M., Mathur, M., Soderberg, C. K., Denis, A., Perfito, N., Iorns, E., & Nosek, B. A. (2021). Investigating the replicability of preclinical cancer biology. eLife, 10. https://doi.org/10.7554/eLife.71601
- Hedges, L. V., & Schauer, J. M. (2019a). Consistency of effects is important in replication: Rejoinder to Mathur and VanderWeele (2019). Psychological Methods, 24(5), 576–577. https://doi.org/10.1037/met0000237
- Hedges, L. V., & Schauer, J. M. (2019b). More Than One Replication Study Is Needed for Unambiguous Tests of Replication. Journal of Educational and Behavioral Statistics: A Quarterly Publication Sponsored by the American Educational Research Association and the American Statistical Association, 44(5), 543–570. https://doi.org/10.3102/1076998619852953
- Ly, A., Etz, A., Marsman, M., & Wagenmakers, E.-J. (2019). Replication Bayes factors from evidence updating. Behavior Research Methods, 51(6), 2498–2508. https://doi.org/10.3758/s13428-018-1092-x
- Mathur, M. B., & VanderWeele, T. J. (2019). Challenges and suggestions for defining replication “success” when effects may be heterogeneous: Comment on Hedges and Schauer (2019). Psychological Methods, 24(5), 571–575. https://doi.org/10.1037/met0000223
- Mathur, M. B., & VanderWeele, T. J. (2020). New statistical metrics for multisite replication projects. Journal of the Royal Statistical Society. Series A, (Statistics in Society), 183(3), 1145–1166. https://doi.org/10.1111/rssa.12572
- Schauer, J. M., & Hedges, L. V. (2021). Reconsidering statistical methods for assessing replication. Psychological Methods, 26(1), 127–139. https://doi.org/10.1037/met0000302
- Schauer, Jacob M. (2023). On the Accuracy of Replication Failure Rates. Multivariate Behavioral Research, 58(3), 598–615. https://doi.org/10.1080/00273171.2022.2066500
- Schauer, Jacob M., Fitzgerald, K. G., Peko-Spicer, S., Whalen, M. C. R., Zejnullahi, R., & Hedges, L. V. (2021). An evaluation of statistical methods for aggregate patterns of replication failure. The Annals of Applied Statistics, 15(1), 208–229. https://doi.org/10.1214/20-AOAS1387
- Schauer, Jacob M., & Hedges, L. V. (2020). Assessing heterogeneity and power in replications of psychological experiments. Psychological Bulletin, 146(8), 701–719. https://doi.org/10.1037/bul0000232
- Simonsohn, U. (2015). Small telescopes: detectability and the evaluation of replication results. Psychological Science, 26(5), 559–569. https://doi.org/10.1177/0956797614567341
- Spence, J. R., & Stanley, D. J. (2016). Prediction interval: What to expect when you’re expecting … A replication. PloS One, 11(9), e0162874. https://doi.org/10.1371/journal.pone.0162874
- Valentine, J. C., Biglan, A., Boruch, R. F., Castro, F. G., Collins, L. M., Flay, B. R., Kellam, S., Mościcki, E. K., & Schinke, S. P. (2011). Replication in prevention science. Prevention Science: The Official Journal of the Society for Prevention Research, 12(2), 103–117. https://doi.org/10.1007/s11121-011-0217-6
- Nosek, B. A., Hardwicke, T. E., Moshontz, H., Allard, A., Corker, K. S., Dreber, A., Fidler, F., Hilgard, J., Kline Struhl, M., Nuijten, M. B., Rohrer, J. M., Romero, F., Scheel, A. M., Scherer, L. D., Schönbrodt, F. D., & Vazire, S. (2022). Replicability, robustness, and reproducibility in psychological science. Annual Review of Psychology, 73(1), 719–748. https://doi.org/10.1146/annurev-psych-020821-114157
- Fletcher, S. C. (2021). How (not) to measure replication. European Journal for Philosophy of Science, 11(2), 57. https://doi.org/10.1007/s13194-021-00377-2


## Chapter 5 - meta-analysis and publication bias

<!-- TODO check if putting this -->

- quick intro to meta-analysis (for the publication bias)
- publication bias as an important replication problem
- statistical method for the publication bias assessment and correction
- p curve and z curve

#### References

- Bartoš, F., & Schimmack, U. (2022). Z-curve 2.0: Estimating replication rates and discovery rates. Meta-Psychology. https://conferences.lnu.se/index.php/metapsychology/article/view/2720
- Simonsohn, U., Nelson, L. D., & Simmons, J. (2013). P-curve: A key to the file drawer. Journal of Experimental Psychology. General, 143(2), 534. https://doi.org/10.1037/a0033242
- Simonsohn, Uri, Nelson, L. D., & Simmons, J. P. (2014). P-curve and effect size: Correcting for publication bias using only significant results: Correcting for publication bias using only significant results. Perspectives on Psychological Science: A Journal of the Association for Psychological Science, 9(6), 666–681. https://doi.org/10.1177/1745691614553988

## Chapter 6 - improving the studies

<!-- TODO check if putting this -->

- multiverse analysis, exploratory and inference
- vibration analysis
- specification curve
- pima

#### References

- Patel, C. J., Burford, B., & Ioannidis, J. P. A. (2015). Assessment of vibration of effects due to model specification can demonstrate the instability of observational associations. Journal of Clinical Epidemiology, 68(9), 1046–1058. https://doi.org/10.1016/j.jclinepi.2015.05.029
- Girardi, P., Vesely, A., Lakens, D., Altoè, G., Pastore, M., Calcagnì, A., & Finos, L. (2024). Post-selection Inference in Multiverse Analysis (PIMA): An Inferential Framework Based on the Sign Flipping Score Test. Psychometrika. https://doi.org/10.1007/s11336-024-09973-6
- Simonsohn, U., Simmons, J. P., & Nelson, L. D. (2020). Specification curve analysis. Nature Human Behaviour, 4(11), 1208–1214. https://doi.org/10.1038/s41562-020-0912-z


## Chapter 7 - planning replication studies

- hedges model and problems

#### References

- Hedges, L. V., & Schauer, J. M. (2021). The Design of Replication Studies. Journal of the Royal Statistical Society. Series A, , 184(3), 868–886. https://doi.org/10.1111/rssa.12688

```{r, results = "asis"}
system("todoget scan")
todo <- readLines("todoget.md")
cat(todo, sep = "\n")
```